### 2025-05-06 manuscript/manuscript.Rmd
Adding potential deep vs surface labeling comparison figure

parametadata <- read_csv("metadata/parametadata.csv") %>%
  # filter(IS_type=="noIS") %>%
  filter(polarity=="pos") %>%
  filter(samp_type=="Smp") %>%
  mutate(startime=factor(startime, levels=c("Morn", "Eve"))) %>%
  mutate(depth=factor(depth, levels=c("Surf", "Deep"))) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Nit", "Arg", "GMP"))) %>%
  arrange(amendment, depth, timepoint_num, startime) %>%
  mutate(timepoint=fct_inorder(timepoint))
peak_areas <- read_csv("targeted/all_peak_areas_w_isos.csv") %>%
  left_join(parametadata %>% distinct(filename, shortname))
all_stans <- read_csv("metadata/all_stans.csv")
maa_names <- c("Also Palythene/Usujirene?", "Shinorine?", "Mycosporine-glycine?", 
               "Palythine?", "Another Palythene/Usujirene?", "Porphyra-334?", 
               "Also Porphyra-334?", "Asterina-330?", "Palythene/Usujirene?",
               "Mycosporine-2-glycine?")
nucleo_names <- c("5-Methylcytosine", "Adenosine", "Cytidine", "Adenine", "Guanosine",
                  "Cytosine", "Hypoxanthine/Allopurinol", "Guanine", "Deoxycytidine?",
                  "Deoxyadenosine", "Deoxyguanosine?")
betaine_names <- c("beta-Alaninebetaine", "Glycine betaine", "Homarine", "Trigonelline",
                   "Choline", "Glycerophosphocholine", "Homoserine betaine?", 
                   "Threonine betaine?", "Proline betaine",
                   "(3-Carboxypropyl)trimethylammonium", "O-Acetylcarnitine", 
                   "Betonicine", "Acetylcholine", "Phosphocholine")
non_n_names <- c("Dimethylsulfonioacetate", "Dimethylsulfoniopropionate",
                 "Arsenobetaine", "Dimethylsulfoxonium propionate?", "Gonyol",
                 "2-O-alpha-D-Glucosylglycerol")
aa_names <- c("Glutamine", "Aspartic acid", "Glutamic acid/beta-Glutamic acid",
              "Asparagine", "Alanine", "Methionine", "Arginine", "Proline",
              "Lysine", "Threonine/Homoserine", "Serine", "Glycine", "Histidine",
              "Leucine/Isoleucine", "Valine", "Arginine")
bad_cmpds <- c("Melamine", "Uridine", "Itaconic acid?", "Inosine", "Dexpanthenol", 
               "Choline sulfate", "5-Hydroxyectoine", "L-Tyrosine")
samp_areas <- parametadata %>%
  filter(samp_type=="Smp") %>%
  distinct(shortname, depth) %>%
  left_join(peak_areas) %>%
  mutate(shortname=fct_inorder(shortname)) %>%
  filter(!compound_name%in%bad_cmpds) %>%
  filter(!(depth=="Deep" & compound_name%in%maa_names)) %>%
  select(-depth)
labeled_fracs <- samp_areas %>%
  mutate(labeled_frac=1-area/sum(area), .by = c("shortname", "compound_name")) %>%
  select(compound_name, iso_name, shortname, labeled_frac) %>%
  filter(str_detect(iso_name, "13C0, 15N0")) %>%
  select(-iso_name)
n_labeled_fracs <- samp_areas %>%
  filter(str_detect(iso_name, "13C0")) %>%
  filter(!str_detect(iso_name, "15N0") | str_detect(iso_name, "13C0, 15N0")) %>%
  mutate(labeled_frac=1-area/sum(area), .by = c("shortname", "compound_name")) %>%
  select(compound_name, iso_name, shortname, labeled_frac) %>%
  filter(str_detect(iso_name, "13C0, 15N0")) %>%
  select(-iso_name)
log_coef_df <- labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, amendment, depth, timepoint_num)) %>%
  nest(data=c(shortname, timepoint_num, labeled_frac)) %>%
  mutate(log_fit=map(data, ~glm(.x$labeled_frac~.x$timepoint_num, family="quasibinomial")),
         coefs=map(log_fit, function(x)broom::tidy(x))) %>%
  unnest(coefs) %>%
  select(compound_name, amendment, depth, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  rename(intercept=`(Intercept)`, slope=`.x$timepoint_num`) %>%
  mutate(half_max=-intercept/slope, scale=1/slope) %>%
  mutate(frac_at_T24 = 1/(1+exp(-(intercept+slope*24))))
```

@@

>>>>

parametadata <- read_csv("metadata/parametadata.csv") %>%
  # filter(IS_type=="noIS") %>%
  filter(polarity=="pos") %>%
  filter(samp_type=="Smp") %>%
  mutate(startime=factor(startime, levels=c("Morn", "Eve"))) %>%
  mutate(depth=factor(depth, levels=c("Surf", "Deep"))) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Nit", "Arg", "GMP"))) %>%
  arrange(amendment, depth, timepoint_num, startime) %>%
  mutate(timepoint=fct_inorder(timepoint))
peak_areas <- read_csv("targeted/all_peak_areas_w_isos.csv") %>%
  left_join(parametadata %>% distinct(filename, shortname))
all_stans <- read_csv("metadata/all_stans.csv")
maa_names <- c("Also Palythene/Usujirene?", "Shinorine?", "Mycosporine-glycine?", 
               "Palythine?", "Another Palythene/Usujirene?", "Porphyra-334?", 
               "Also Porphyra-334?", "Asterina-330?", "Palythene/Usujirene?",
               "Mycosporine-2-glycine?")
nucleo_names <- c("5-Methylcytosine", "Adenosine", "Cytidine", "Adenine", "Guanosine",
                  "Cytosine", "Hypoxanthine/Allopurinol", "Guanine", "Deoxycytidine?",
                  "Deoxyadenosine", "Deoxyguanosine?")
betaine_names <- c("beta-Alaninebetaine", "Glycine betaine", "Homarine", "Trigonelline",
                   "Choline", "Glycerophosphocholine", "Homoserine betaine?", 
                   "Threonine betaine?", "Proline betaine",
                   "(3-Carboxypropyl)trimethylammonium", "O-Acetylcarnitine", 
                   "Betonicine", "Acetylcholine", "Phosphocholine")
non_n_names <- c("Dimethylsulfonioacetate", "Dimethylsulfoniopropionate",
                 "Arsenobetaine", "Dimethylsulfoxonium propionate?", "Gonyol",
                 "2-O-alpha-D-Glucosylglycerol")
aa_names <- c("Glutamine", "Aspartic acid", "Glutamic acid/beta-Glutamic acid",
              "Asparagine", "Alanine", "Methionine", "Arginine", "Proline",
              "Lysine", "Threonine/Homoserine", "Serine", "Glycine", "Histidine",
              "Leucine/Isoleucine", "Valine", "Arginine")
bad_cmpds <- c("Melamine", "Uridine", "Itaconic acid?", "Inosine", "Dexpanthenol", 
               "Choline sulfate", "5-Hydroxyectoine", "L-Tyrosine")
samp_areas <- parametadata %>%
  filter(samp_type=="Smp") %>%
  distinct(shortname, depth) %>%
  left_join(peak_areas) %>%
  mutate(shortname=fct_inorder(shortname)) %>%
  filter(!compound_name%in%bad_cmpds) %>%
  filter(!(depth=="Deep" & compound_name%in%maa_names)) %>%
  select(-depth)
labeled_fracs <- samp_areas %>%
  mutate(labeled_frac=1-area/sum(area), .by = c("shortname", "compound_name")) %>%
  select(compound_name, iso_name, shortname, labeled_frac) %>%
  filter(str_detect(iso_name, "13C0, 15N0")) %>%
  select(-iso_name)
log_coef_df <- labeled_fracs %>%
  filter(!compound_name%in%maa_names) %>%
  left_join(parametadata %>% distinct(shortname, amendment, depth, timepoint_num)) %>%
  nest(data=c(shortname, timepoint_num, labeled_frac)) %>%
  mutate(log_fit=map(data, ~glm(.x$labeled_frac~.x$timepoint_num, family="quasibinomial")),
         coefs=map(log_fit, function(x)broom::tidy(x))) %>%
  unnest(coefs) %>%
  select(compound_name, amendment, depth, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  rename(intercept=`(Intercept)`, slope=`.x$timepoint_num`) %>%
  mutate(half_max=-intercept/slope, scale=1/slope) %>%
  mutate(frac_at_T24 = 1/(1+exp(-(intercept+slope*24))))
```

@@

```{r fig surface vs deep labeling rates}
labeled_fracs %>%
  filter(!compound_name%in%maa_names) %>%
  left_join(distinct(parametadata, shortname, amendment, depth, timepoint_num)) %>%
  filter(amendment!="Nit") %>%
  summarise(avg_label=mean(labeled_frac), .by=c(compound_name, amendment, depth, timepoint_num)) %>%
  pivot_wider(names_from = depth, values_from=avg_label) %>%
  mutate(timepoint_num=ifelse(timepoint_num==0, 0.3, timepoint_num)) %>%
  ggplot(aes(x=timepoint_num, y=Surf-Deep, color=compound_name, group=compound_name)) +
  geom_hline(yintercept = 0) +
  geom_line() +
  geom_point() +
  facet_wrap(~amendment, nrow=1) +
  scale_x_log10() +
  theme_bw() +
  theme(legend.position = "none")
ggsave(filename = "deep_vs_surf_labeling.png", device = "png", path = "manuscript/figures",
       width = 6.5, height = 4, units = "in", dpi = 300)
```

![](figures/deep_vs_surf_labeling.png)

*Figure XX: Line plots of the metabolite fraction labeled at the surface (25 meters) minus the fraction labeled at 175 meters. Incubations with added ammonia, arginine, and guanosine monophosphate (GMP) are shown. Points are the average of six biological replicates (incubations started in the morning and the evening are combined for additional power) and colors have been added solely for disambiguating purposes. Positive values indicate increased synthesis in the surface samples relative to those at depth.*

```{r regression attempts currently failing, eval=FALSE}
labeled_fracs %>%
  left_join(parametadata %>% filter(IS_type=="noIS")) %>%
  filter(compound_name=="Creatinine") %>%
  filter(amendment=="GMP") %>%
  mutate(timepoint_num=ifelse(timepoint_num==0, 0.3, timepoint_num)) %>%
  ggplot(aes(x=timepoint_num, y=labeled_frac, color=startime)) +
  geom_point() +
  geom_smooth(method=glm, formula=y~x, method.args=list(family=quasibinomial)) +
  facet_wrap(~depth, ncol=1)

reg_data <- labeled_fracs %>%
  filter(!compound_name%in%maa_names) %>%
  left_join(distinct(parametadata, shortname, amendment, depth, timepoint_num)) %>%
  mutate(max_frac=max(labeled_frac[timepoint_num==max(timepoint_num)]), .by=c(amendment, compound_name, depth)) %>%
  mutate(t10_labfrac = mean(labeled_frac[timepoint_num==sort(unique(timepoint_num))[4]]), .by=c(amendment, compound_name, depth))

reg_data %>%
  distinct(compound_name, amendment, depth, max_frac, t10_labfrac) %>%
  mutate(category=ifelse(max_frac < 0.1, "Unlabeled", "Good")) %>%
  mutate(category=ifelse(t10_labfrac <= 0, "Only T26 labeled", category)) %>%
  mutate(category=ifelse(is.na(category), "DMSOP (unlabeled)", category)) %>%
  count(amendment, category, depth) %>%
  print(n=Inf)

reg_fits <- reg_data %>%
  filter(max_frac>0.1) %>%
  filter(t10_labfrac>0) %>%
  filter(compound_name!="Betonicine") %>%
  filter(compound_name!="Arsenobetaine") %>%
  filter(compound_name!="Gonyol") %>%
  mutate(timepoint_num=ifelse(timepoint_num==0, 0.3, timepoint_num)) %>%
  nest(data=c(shortname, timepoint_num, labeled_frac)) %>%
  mutate(log_fit=map(data, ~glm(.x$labeled_frac~.x$timepoint_num, family="quasibinomial")),
         coefs=map(log_fit, function(x)broom::tidy(x))) %>%
  unnest(coefs) %>%
  mutate(p_adj=p.adjust(p.value, method = "fdr"), .by = term) %>%
  filter(p_adj<0.001) %>%
  select(compound_name, amendment, depth, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  rename(intercept=`(Intercept)`, slope=`.x$timepoint_num`) %>%
  mutate(half_max=-intercept/slope, scale=1/slope) %>%
  mutate(frac_at_T24 = 1/(1+exp(-(intercept+slope*24))))
# reg_fits %>%
#   select(compound_name, amendment, depth, intercept, slope) %>%
#   pivot_longer(intercept:slope) %>%
#   ggplot(aes(x=depth, y=value)) +
#   geom_boxplot() +
#   facet_grid(name~amendment, scales="free_y")
# reg_fits %>%
#   select(compound_name, amendment, depth, intercept, slope) %>%
#   pivot_longer(intercept:slope) %>%
#   pivot_wider(names_from=depth) %>%
#   filter(!is.na(Deep)) %>%
#   nest(data=c(compound_name, Surf, Deep)) %>%
#   mutate(test=map(data, ~broom::tidy(t.test(.x$Surf, .x$Deep, paired=TRUE)))) %>%
#   unnest(test)
# labeled_fracs %>%
#   left_join(parametadata %>% distinct(shortname, amendment, depth, timepoint_num)) %>%
#   filter(timepoint_num==max(timepoint_num), .by=amendment) %>%
#   summarise(avg_Tf_label=mean(labeled_frac), .by=c(compound_name, amendment, depth)) %>%
#   ggplot() +
#   geom_boxplot(aes(x=amendment, y=avg_Tf_label, color=depth))
# 
# labeled_fracs %>%
#   left_join(parametadata %>% distinct(shortname, amendment, depth, timepoint_num)) %>%
#   filter(timepoint_num==max(timepoint_num), .by=amendment) %>%
#   summarise(avg_Tf_label=mean(labeled_frac), .by=c(compound_name, amendment, depth)) %>%
#   pivot_wider(names_from=depth, values_from=avg_Tf_label) %>%
#   ggplot() +
#   geom_segment(aes(y=Surf, yend=Deep), x=1, xend=2) +
#   facet_wrap(~amendment, nrow=1)
# reg_fits %>%
#   ggplot() +
#   geom_segment(aes(y=frac_at_T24_Surf, yend=frac_at_T24_Deep,
#                    x=half_max_Surf, xend=half_max_Deep, color=compound_name),
#                data = reg_fits %>%
#                  select(compound_name, depth, half_max, frac_at_T24) %>% 
#                  pivot_wider(names_from = depth, values_from=c(half_max, frac_at_T24)) %>%
#                  filter(!is.na(half_max_Deep))) +
#   geom_point(aes(x=half_max, y=frac_at_T24, fill=depth), pch=21) +
#   scale_fill_manual(values=c("black", "white"), breaks=c("Deep", "Surf"))

reg_fits %>%
  filter(amendment!="Nit") %>%
  ggplot() +
  geom_segment(aes(y=frac_at_T24_Surf, yend=frac_at_T24_Deep,
                   x=half_max_Surf, xend=half_max_Deep, color=compound_name),
               data = reg_fits %>%
                 select(compound_name, amendment, depth, half_max, frac_at_T24) %>% 
                 pivot_wider(names_from = depth, values_from=c(half_max, frac_at_T24)) %>%
                 filter(!is.na(half_max_Deep))) +
  geom_point(aes(x=half_max, y=frac_at_T24, fill=depth, label=compound_name), pch=21) +
  scale_fill_manual(values=c("black", "white"), breaks=c("Deep", "Surf")) +
  facet_wrap(~amendment, scales="free_x") +
  theme(legend.position = "none")
```
